---
title: "SIT741 - s22407595 - R Notebook"
output: html_notebook
---

#Q.1
# The data preparation including merging the datasets with the time range for both weather and card_accidents data is being implemenbted until line number 380.
# First step involves using the same code from assignment to filter out accident data

# Step 2 involves, preparing the weather data for the required date range
# Step 3 involves, merging data on date and removing unwanted columns
```{r}
# 2.1
library(tidyverse)
library(skimr)

cav_data_link <- 'car_accidents_victoria.csv'
top_row <- read_csv(cav_data_link, col_names = FALSE, n_max = 1)
second_row <- read_csv(cav_data_link, n_max = 1)

column_names <- second_row %>% 
  unlist(., use.names=FALSE) %>% 
  make.unique(., sep = "__") # double underscore

column_names[2:5] <- str_c(column_names[2:5], '0', sep='__')

daily_accidents <-read_csv(cav_data_link, skip = 2, col_names = column_names)

# pivot_longer
daily_accidents1 <- daily_accidents %>%
  pivot_longer(names_to = "ACCIDENT_TYPE_AND_REGION",
              values_to = "NUMBER_OF_ACCIDENTS",
              contains("__"),
              values_drop_na = TRUE
              )

head(daily_accidents1)

# split the columns 
daily_accidents2 <- daily_accidents1 %>% 
  separate(col = ACCIDENT_TYPE_AND_REGION, 
           into = c("ACCIDENT_TYPE", "REGION"), 
           sep = "__",
           extra = "merge")

head(daily_accidents2)

# pivot wider
daily_accidents3 = daily_accidents2 %>% 
  pivot_wider(names_from = ACCIDENT_TYPE, values_from = NUMBER_OF_ACCIDENTS)

head(daily_accidents3)

# convert numeric values to actual values for the region column 
for (value in unique(daily_accidents3$REGION))
{
  region_name <- top_row[!is.na(top_row)][as.numeric(value) + 1]  # +1 since the REGION values starts from 0 and indices in R starts from 1 
  daily_accidents3$REGION[daily_accidents3$REGION == value] <- region_name
}

head(daily_accidents3)

# change to date format
library(lubridate)
daily_accidents3$DATE <- dmy(daily_accidents3$DATE)

# replace missing values with the mode value 
mode_value <- as.numeric(names(sort(table(unlist(daily_accidents3), useNA = "always"), decreasing = TRUE)[1]))

# Replace missing values with the mode
daily_accidents3[is.na(daily_accidents3)] <- mode_value

head(daily_accidents3)

# select only EASTERN region
daily_accidents_SOUTH_EAST_REGION <- daily_accidents3[daily_accidents3$REGION == 'METROPOLITAN SOUTH EAST REGION',]

# compute the total accidents column 
daily_accidents_SOUTH_EAST_REGION <- daily_accidents_SOUTH_EAST_REGION %>%
  mutate(TOTAL_ACCIDENTS = FATAL + SERIOUS + NOINJURY + OTHER)

head(daily_accidents_SOUTH_EAST_REGION)

skim(daily_accidents_SOUTH_EAST_REGION)


daily_accidents_SOUTH_EAST_REGION <- daily_accidents3[daily_accidents3$REGION == 'METROPOLITAN SOUTH EAST REGION',]

daily_accidents_SOUTH_EAST_REGION <- daily_accidents_SOUTH_EAST_REGION %>%
  mutate(TOTAL_ACCIDENTS = FATAL + SERIOUS + NOINJURY + OTHER)
# View(daily_accidents_SOUTH_EAST_REGION)

```

# Prepare/fetch the weather data from NOAA using API calls
```{r}
options(noaakey = "HaxkzDUmmhWPDXwogSALKUUZduhWMHQi")

library(rnoaa)
library(tidyverse)
```


# Prepare 1st data chunk
```{r}
melbourne_data_PRCP <- ncdc(
  datasetid = 'GHCND',
  stationid = 'GHCND:ASN00086077',
  datatypeid = 'PRCP',
  startdate = '2016-01-01',
  enddate = '2016-12-31',
  limit = 1000
)$data

melbourne_data_TMIN <- ncdc(
  datasetid = 'GHCND',
  stationid = 'GHCND:ASN00086077',
  datatypeid = 'TMIN',
  startdate = '2016-01-01',
  enddate = '2016-12-31',
  limit = 1000
)$data

melbourne_data_TMAX <- ncdc(
  datasetid = 'GHCND',
  stationid = 'GHCND:ASN00086077',
  datatypeid = 'TMAX',
  startdate = '2016-01-01',
  enddate = '2016-12-31',
  limit = 1000
)$data

```

```{r}
#merge data
vic_data = rbind(melbourne_data_PRCP, melbourne_data_TMIN)
vic_data = rbind(vic_data, melbourne_data_TMAX)
```


# Prepare 2nd data chunk
```{r}
melbourne_data_PRCP2 <- ncdc(
  datasetid = 'GHCND',
  stationid = 'GHCND:ASN00086077',
  datatypeid = 'PRCP',
  startdate = '2017-01-01',
  enddate = '2017-12-31',
  limit = 1000
)$data

melbourne_data_TMIN2 <- ncdc(
  datasetid = 'GHCND',
  stationid = 'GHCND:ASN00086077',
  datatypeid = 'TMIN',
  startdate = '2017-01-01',
  enddate = '2017-12-31',
  limit = 1000
)$data

melbourne_data_TMAX2 <- ncdc(
  datasetid = 'GHCND',
  stationid = 'GHCND:ASN00086077',
  datatypeid = 'TMAX',
  startdate = '2017-01-01',
  enddate = '2017-12-31',
  limit = 1000
)$data


```

```{r}
# merge the second chunk
vic_data2 = rbind(melbourne_data_PRCP2, melbourne_data_TMIN2)
vic_data2 = rbind(vic_data2, melbourne_data_TMAX2)

```


# Prepare 3rd chunk
```{r}
melbourne_data_PRCP3 <- ncdc(
  datasetid = 'GHCND',
  stationid = 'GHCND:ASN00086077',
  datatypeid = 'PRCP',
  startdate = '2018-01-01',
  enddate = '2018-12-31',
  limit = 1000
)$data

melbourne_data_TMIN3 <- ncdc(
  datasetid = 'GHCND',
  stationid = 'GHCND:ASN00086077',
  datatypeid = 'TMIN',
  startdate = '2018-01-01',
  enddate = '2018-12-31',
  limit = 1000
)$data

melbourne_data_TMAX3 <- ncdc(
  datasetid = 'GHCND',
  stationid = 'GHCND:ASN00086077',
  datatypeid = 'TMAX',
  startdate = '2018-01-01',
  enddate = '2018-12-31',
  limit = 1000
)$data
```

```{r}
#merge data
vic_data3 = rbind(melbourne_data_PRCP3, melbourne_data_TMIN3)
vic_data3 = rbind(vic_data3, melbourne_data_TMAX3)
```

# Prepare 4th chunk
```{r}
melbourne_data_PRCP4 <- ncdc(
  datasetid = 'GHCND',
  stationid = 'GHCND:ASN00086077',
  datatypeid = 'PRCP',
  startdate = '2019-01-01',
  enddate = '2019-12-31',
  limit = 1000
)$data

melbourne_data_TMIN4 <- ncdc(
  datasetid = 'GHCND',
  stationid = 'GHCND:ASN00086077',
  datatypeid = 'TMIN',
  startdate = '2019-01-01',
  enddate = '2019-12-31',
  limit = 1000
)$data

melbourne_data_TMAX4 <- ncdc(
  datasetid = 'GHCND',
  stationid = 'GHCND:ASN00086077',
  datatypeid = 'TMAX',
  startdate = '2019-01-01',
  enddate = '2019-12-31',
  limit = 1000
)$data
```

```{r}
#merge data
vic_data4 = rbind(melbourne_data_PRCP4, melbourne_data_TMIN4)
vic_data4 = rbind(vic_data4, melbourne_data_TMAX4)
```


# Prepare 5th chunk
```{r}
melbourne_data_PRCP5 <- ncdc(
  datasetid = 'GHCND',
  stationid = 'GHCND:ASN00086077',
  datatypeid = 'PRCP',
  startdate = '2020-01-01',
  enddate = '2020-06-30',
  limit = 1000
)$data

melbourne_data_TMIN5 <- ncdc(
  datasetid = 'GHCND',
  stationid = 'GHCND:ASN00086077',
  datatypeid = 'TMIN',
  startdate = '2020-01-01',
  enddate = '2020-06-30',
  limit = 1000
)$data

melbourne_data_TMAX5 <- ncdc(
  datasetid = 'GHCND',
  stationid = 'GHCND:ASN00086077',
  datatypeid = 'TMAX',
  startdate = '2020-01-01',
  enddate = '2020-06-30',
  limit = 1000
)$data
```

```{r}
#merge data
vic_data5 = rbind(melbourne_data_PRCP5, melbourne_data_TMIN5)
vic_data5 = rbind(vic_data5, melbourne_data_TMAX5)
```

# combine all data using rowbind

```{r}
# Row bind the datasets
combined_data <- rbind(vic_data, vic_data2, vic_data3, vic_data4, vic_data5)
#View(combined_data)
rows_count <- nrow(combined_data)
col_count <- ncol(combined_data)

cat("Rows", rows_count, "Cols", col_count)
```

# Merge the data on date
```{r}
# Rename the 'date' column to 'DATE'
names(combined_data)[names(combined_data) == "date"] <- "DATE"

# # Convert the date column to Date type
combined_data <- combined_data %>%
 mutate(DATE = as.Date(DATE))

combined_data <- merge(combined_data, daily_accidents_SOUTH_EAST_REGION, by = "DATE")


combined_data <- combined_data %>%
  spread(key = datatype, value = value)

# Print the modified dataset
print(combined_data)
```
# Replace null values using mode just like from assignment 1
```{r}
library(dplyr)
library(stats)

# Define a function to calculate the mode, ignoring NA values
calculate_mode <- function(x) {
  uniq_x <- unique(x)
  uniq_x[which.max(tabulate(match(x, uniq_x)))]
}

# Replace missing values with the mode for each column (excluding the date column)
combined_data <- combined_data %>%
  mutate(across(where(~ !is.Date(.)), ~ ifelse(is.na(.), calculate_mode(.), .)))

# Print the updated dataset
print(combined_data)


which(is.na(combined_data), arr.ind = TRUE)
```

```{r}
library(dplyr)

final_data <- combined_data %>%
  select(-station, -fl_m, -fl_q, -fl_so, -fl_t, -FATAL, -NOINJURY, -SERIOUS, -OTHER)


final_data

```

```{r}

null_counts <- sapply(final_data, function(col) sum(is.na(col)))
print(null_counts)

tot_rows <- nrow(final_data)
cat("Tot_rows: ", tot_rows, '\n')

tot_cols <- ncol(final_data)
cat('Tot Cols', tot_cols, '\n')
# combined_data <- combined_data %>% mutate(date = dmy(date))

data_types <-sapply(final_data, class)
cat('Data tyes', data_types)

# Find the minimum and maximum dates in the dataset
max_date = max(final_data$date)
max_date <- max(final_data$DATE, na.rm = TRUE)
min_date <- min(final_data$DATE, na.rm = TRUE)

sprintf("Time range of data is from %s to %s", min_date, max_date)
```




# Question 3
# Southeast region was chosen, using the assignment 1 code for accident data and have hand picked the data from NOAA for southeast region data
# Q3.2
```{r}
# Load libraries
library(ggplot2)
library(dplyr)

# Use a linear model
lm_model <- lm(TOTAL_ACCIDENTS ~ DATE, data = final_data)

# Create a dataframe with date and fitted values
fitted_data <- final_data %>%
  select(DATE) %>%
  mutate(fitted_values = predict(lm_model))

p1 <- ggplot(final_data, aes(x = DATE, y = TOTAL_ACCIDENTS)) +
  geom_point() +
  geom_line(data = fitted_data, aes(x = DATE, y = fitted_values), color = "blue") +
  labs(title = "Fitted Values vs. Date", x = "Date", y = "TOTAL_ACCIDENTS")

# Create a plot for residuals
p2 <- ggplot(final_data, aes(x = DATE, y = residuals(lm_model))) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red") +
  labs(title = "Residuals vs. Date", x = "Date", y = "Residuals")

final_data <- final_data %>%
  mutate(standardized_residuals = residuals(lm_model) / sd(residuals(lm_model)))


# Create a plot for standardized residuals
p3 <- ggplot(final_data, aes(x = DATE, y = standardized_residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red") +
  labs(title = "Standardized Residuals vs. Date", x = "Date", y = "Standardized Residuals")


# Summary
summary(lm_model)

```


```{r}
# Print Fitted values vs date
print(p1)
```

```{r}
#Residuals vs Date plot
print(p2)

```
```{r}
# Standardized residulas plot
print(p3)
```


# Try QQ plot
```{r}

#try the QQ-plot for residuals to see whether they follow a theoretical distribution
library(dplyr)
lm_model %>% 
  ggplot(aes(sample = .resid)) +
  geom_qq() + 
  geom_qq_line(col = 'blue') + 
  xlab('Theoritical Quantiles') + 
  ylab('Standardized residuals')

```


# Q3.3
# GAM model
```{r}
# Load necessary libraries
library(mgcv)
library(ggplot2)


final_data <- mutate(final_data, date_numeric = as.numeric(DATE))

gam_model <- gam(TOTAL_ACCIDENTS ~ s(date_numeric), data = final_data)

# Summary of the GAM Model Fit
summary(gam_model)
```
```{r}
# Plot the GAM fit
plot(gam_model, rug = TRUE, residuals = TRUE, pch = 1, cex = 1, shade = TRUE, shade.col = "yellow")

```
```{r}
gam.check(gam_model)

```


```{r}
gam.check(gam_model)

```
# Augment the model
# 3.4
```{r}
#Adding week day to the dataset
library(dplyr)    # For data manipulation
library(lubridate) # For working with dates
library(mgcv)

week_wise_data = final_data %>%
  mutate(week_day = wday(DATE, label = TRUE))


#Converting Week Days into numeric
week_wise_data = week_wise_data %>%
  mutate(week_day_numeric = as.numeric(week_day))

weekly_gam_model = gam(TOTAL_ACCIDENTS ~ s(date_numeric) + s(week_day_numeric, k = 7),
                        data = week_wise_data)
# summary
weekly_gam_model %>% summary
```
```{r}
gam.check(weekly_gam_model)


```



```{r}
plot(weekly_gam_model, rug = TRUE, residuals = TRUE, pch = 1, cex = 1, shade = TRUE, shade.col = "Yellow")

```


# Q3.5 - AIC scores
```{r}
print("AIC score of Linear Model:")
AIC(lm_model)
print("AIC score of GAM Model:")
AIC(gam_model)
print("AIC score of GAM Model with weekly variance:")
AIC(weekly_gam_model)
```


# Question 4.1
```{r}
library(rnoaa)
library(tidyverse)
final_data_30years_TMAX = data.frame(date = character(),
                              datatype = character(),
                              station = character(),
                              value = integer(),
                              fl_m = character(),
                              fl_q = character(),
                              fl_so = character(),
                              fl_t = character(),
                              stringsAsFactors = FALSE)

# get for 30 years
for(i in 1:30) { 

  year = 1970 + i
  startDate = paste(year, '-01-01', sep = '')
  endDate = paste(year, '-12-31', sep = '')
  
  yearlyData = ncdc(
    datasetid = 'GHCND',
    stationid = 'GHCND:ASN00086077',
    datatypeid = 'TMAX',
    startdate = startDate,
    enddate = endDate,
    limit = 1000
  )$data
  
  final_data_30years_TMAX = rbind(final_data_30years_TMAX, yearlyData)
}

```

```{r}

final_data_30years_TMIN = data.frame(date = character(),
                              datatype = character(),
                              station = character(),
                              value = integer(),
                              fl_m = character(),
                              fl_q = character(),
                              fl_so = character(),
                              fl_t = character(),
                              stringsAsFactors = FALSE)

# get for 30 years
for(i in 1:30) { 
  
  year = 1970 + i
  startDate = paste(year, '-01-01', sep = '')
  endDate = paste(year, '-12-31', sep = '')
  
  yearlyData = ncdc(
    datasetid = 'GHCND',
    stationid = 'GHCND:ASN00086077',
    datatypeid = 'TMIN',
    startdate = startDate,
    enddate = endDate,
    limit = 1000
  )$data
  
  final_data_30years_TMIN = rbind(final_data_30years_TMIN, yearlyData)
}

```

```{r}
final_data_30years_TMIN_updated <- final_data_30years_TMIN %>%
  select(-datatype, -station, -fl_m, -fl_q, -fl_so, -fl_t)
colnames(final_data_30years_TMIN_updated)[2] = "TMIN"

final_data_30years_TMAX_updated <- final_data_30years_TMAX %>%
  select(-datatype, -station, -fl_m, -fl_q, -fl_so, -fl_t)
colnames(final_data_30years_TMAX_updated)[2] = "TMAX"

#Merging TMAX and TMIN updated datasets
final_data_30years_merged = merge(x = final_data_30years_TMIN_updated, y = final_data_30years_TMAX_updated, 
                        by = "date", all.x = TRUE)

#Calculating DMT of 30 years data
final_data_30years_merged$DMT = rowMeans(final_data_30years_merged[, c('TMAX', 'TMIN')], na.rm = TRUE)

# Function to calculate mode
calculate_mode <- function(x) {
  uniq_x <- unique(x)
  uniq_x[which.max(tabulate(match(x, uniq_x)))]
}

# Replace missing values in the TMIN column with mode
final_data_30years_merged <- final_data_30years_merged %>%
  mutate(TMIN = ifelse(is.na(TMIN), calculate_mode(TMIN), TMIN))


# calculate DMT for the final_data
final_data$DMT = rowMeans(final_data[, c('TMAX', 'TMIN')], na.rm = TRUE)


#Calculating DMT of 30 years data
final_data_30years_merged$DMT = rowMeans(final_data_30years_merged[, c('TMAX', 'TMIN')], na.rm = TRUE)

```

```{r}
t95 = quantile(final_data_30years_merged$DMT, 0.95) 
print(t95)
```

```{r}
# install the package zoo to analyze time-series data

# find the 3 days average using width=3
final_data$ThreeDaysAVG = zoo::rollapply(final_data$DMT, width = 3, FUN = mean,
                                        fill = NA, align = 'left')

```


```{r}
#Calculating EHI sig.
final_data$EHIsig = final_data$ThreeDaysAVG - t95
```


```{r}
#Finding 30 days average of DMT
final_data$ThirtyDaysAVG = zoo::rollapplyr(c(rep(NA, 1), final_data$DMT[1:(length(final_data$DMT)-1)]), width = 30,FUN = mean, fill = NA)
```


```{r}
#calculate EHI Acclimatization
final_data$EHIaccl = final_data$ThreeDaysAVG - final_data$ThirtyDaysAVG
```

```{r}
#calculate the EHF (Excess Heat Factor)
# ð¸ð»ð¹=ð¸ð»ð¼sigÃ—max(1, ð¸ð»ð¼accl)
# This formula is obtained from the paper


# final_data$EHF = pmax(0, final_data$EHIsig) * pmax(1, final_data$EHIaccl)
# use this code below to get the periodical output
final_data$EHF = final_data$EHIsig * pmax(1, final_data$EHIaccl)
final_data = final_data %>% mutate(EHF = ifelse(is.na(EHF), 0, EHF))
```

```{r}
#Plotting the EHF value
library(ggplot2)

# Convert date column to POSIXct
final_data <- final_data %>%
  mutate(date = as.POSIXct(DATE, format = "%Y-%m-%d"))

# Check the structure of the date column
str(final_data$DATE)

# Create the plot using ggplot
ggplot(final_data, aes(x = date, y = EHF, group = 1)) +
  geom_line(color = 'blue') +
  ylab("Excess Heat Factor")
```


```{r}

#Building GAM model
# Use EHF as additional factor

ehf_gam_model = gam(TOTAL_ACCIDENTS ~ s(date_numeric) + s(week_day_numeric, k = 7) + 
                  s(final_data$EHF), data = week_wise_data)
summary(ehf_gam_model)

```

```{r}
plot(ehf_gam_model, rug = TRUE, residuals = TRUE, pch = 1, cex = 1,
     shade = TRUE, shade.col = "blue")
```

```{r}
gam.check(ehf_gam_model)
```

```{r}
print("AIC score of GAM Model with EHF is:")
AIC(ehf_gam_model)
```


Task 4.3 Model with extra features
```{r}
#Building GAM Fit Model for Attendance with EHF and another feature (Temperature range)

final_data$TRange <- final_data$TMAX - final_data$TMIN

extra_pred_gam_trange = gam(TOTAL_ACCIDENTS ~ s(date_numeric) + s(week_day_numeric, k = 7)
                  + s(final_data$EHF) + s(final_data$TRange), data = week_wise_data)

plot(extra_pred_gam_trange, rug = TRUE, residuals = TRUE, pch = 1, cex = 1,
     shade = TRUE, shade.col = "blue")

gam.check(extra_pred_gam_trange)
```

```{r}
summary(extra_pred_gam_trange)
```

```{r}
print("AIC score of GAM Model with EHF and Temperature range is:")
AIC(extra_pred_gam_trange)

```